On Optimality of Finite Element Integration

### 2015-09-21 TOMS_Paper/assembly.tex
Rework section 2 entirely...

\section{Optimality of Loop Nests}
\label{sec:optimal-impl}

In this section, we provide a characterization of loop nest optimality. In order to make the document self-contained, we start with reviewing basic compiler terminology.

@@

By mapping mathematical properties to the loop nest level, we introduce the
concept of a \textit{linear loop}.

@@

In particular, we are interested in the following class since it naturally arises in finite element integration.

\begin{Def}[Perfect multilinear loop nest]
A \textit{perfect multilinear} loop nest of arity $n$ is a perfect nest composed of $n$ loops, in which all of the expressions appearing in the body of the innermost loop are linear in each loop $L_i$ separately.
\end{Def}

Note that perfect multilinear loop nests could actually be rooted in deeper, possibly imperfect loop nests. In fact, we will focus on this particular structure.

We introduce the fundamental notion of sharing.

@@

Figure~\ref{code:multi_loopnest_a} shows an example of a trivial multilinear loop nest of arity $n=2$ with sharing along dimension $j$.

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 7pt \subfigure[With sharing]{\label{code:multi_loopnest_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}}}
~~~~~~~~~~~~~~
{\subfigcapskip = 2pt \subfigure[Optimal form]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}}}
\end{CenteredBox}\caption{A simple multilinear loop nest}\end{figure}

We now have the ingredients to formulate a simple yet fundamental result.

\begin{Prop}
\label{prop:share-removal}
Assume $\Lambda = [L_{i_{0}}, L_{i_{1}}, ..., L_{i_{n-1}}]$ is a perfect multilinear loop nest with the body of $L_{i_{n-1}}$ being an independent basic block. Then sharing can always be eliminated from $\Lambda$.
\end{Prop}
\begin{proof}
  The demonstration is by construction and exploits linearity. We want to
  transform $\Lambda$ into $\Lambda'$ such that there is no sharing in any
  $L_i \in \Lambda'$. Starting from the innermost loop $L_{i_{n-1}}$, the
  expressions are ``flattened'' by expanding all products involving terms
  depending on $L_{i_{n-1}}$. Being now on the same level of the expression
  tree, such terms can be factorized. Due to linearity, each factored
  product only has one term depending on $L_{i_{n-1}}$, and this term is now
  unique in the expression. The other terms, independent of $L_{i_{n-1}}$,
  are, by definition, loop-invariant, and as such can be hoisted at the
  level of $L_{i_{n-2}}$. This procedure can be applied recursively up to
  $L_{i_{0}}$: multilinearity allows factorization at each level;
  perfectness ensures hoisting is safe.
\end{proof}

This result will be essential to prove that optimality, characterized as follows, can always be reached in this class of loops.

\begin{Def}[Optimality of a multilinear loop nest]
\label{def:mln-optimality}
The synthesis of a multilinear loop nest is optimal if the number of
operations performed in the loop nest is minimal.
\end{Def}

%We will see that the restriction to inner loops has the effect of simplifying the transformation model; the price to pay will be the generation of quasi-optimal code in presence of costly outer loops (this, in finite element integration, happens when forms are characterized by extensive use of tensor algebra).

Note that Definition~\ref{def:mln-optimality} does not take into account
memory requirements. If the loop nest were memory-bound -- the ratio of
operations to bytes transferred from memory to the CPU being too low -- then
speaking of optimality would clearly make no sense. In the following, we
assume to operate in a CPU-bound regime, in which arithmetic-intensive
expressions need be evaluated. In the context of finite element, this is
often true for more complex multilinear forms and/or higher order elements.

%\begin{Rem*}
%In the following, when discussing ``optimality'' or ``loop nests in optimal form'', we will always tacitly refer to inner optimality (unless explicitly specified differently). 
%\end{Rem*}

As anticipated, perfect multilinear loop nests can be brought to optimal form. The following result proves this claim.

\begin{Prop}
\label{prop:optimal-mln}
A perfect multilinear loop nest $\Lambda$ with basic block independence can always be reduced to optimal form.
\end{Prop}
\begin{proof}
By construction. Loop-dependent terms are logically grouped into $n$
disjoint sets $S_i$, each $S_i$ containing all terms depending on
$L_i$. These sets are sorted in descending order based on their
cardinality. We produce a new loop permutation that reflects the ${S_i}$
ordering; that is, loops are placed such that each loop has fewer unique
terms than the outer loops containing it. Since the loop nest is perfect,
any such permutation is semantically correct.  Starting from $L_{i_{n-1}}$, we can apply the sharing-removal procedure described in Proposition~\ref{prop:share-removal}. This renders $L_{i_{n-1}}$ optimal. In particular, the number of operations is equal to $\Lambda_{ops} = \# S_{i_{n-1}} + (\# S_{i_{n-1}} - 1)$, in which the second term represents the cost of summing the $\# S_{i_{n-1}}$ terms.
\end{proof}

For example, by applying the construction described in Proposition~\ref{prop:optimal-mln} to the code in Figure~\ref{code:multi_loopnest_a}, we obtain the optimal form in Figure~\ref{code:multi_loopnest_b}.
\\
\begin{figure}[h]\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest.code}
\end{CenteredBox}\caption{An imperfect loop nest, with one reduction loop ($L_i$), an inner perfect multilinear loop nest ($[L_j, L_k]$), and sharing potentially present in $F$}\label{code:loopnest}\end{figure}

We now consider a wider class of loop nests in which perfectness and multilinearity apply to a sub-nest only. Consider the example in Figure~\ref{code:loopnest}. The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises a reduction loop $L_i$ and a perfect doubly nested loop $[L_j, L_k]$, which we assume to be multilinear. The right hand side of the statement computing the multidimensional array \texttt{A} is a generic expression \texttt{F} including standard arithmetic operations such as addition and multiplication, possibly characterized by sharing. Here, one could think of searching for sub-expressions for which the reduction could be pre-evaluated, thus obtaining a decrease proportional to $M$ in the operation count. This is exemplified through Figures~\ref{code:loopnest_red} and~\ref{code:loopnest_nored} by fixing $F$ in a simpler loop nest.

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 19pt \subfigure[With reduction]{\label{code:loopnest_red}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_red.code}}}
~~~~~~~~~~
{\subfigcapskip = 2pt \subfigure[Pre-evaluated reduction]{\label{code:loopnest_nored}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_nored.code}}}
\end{CenteredBox}\caption{The effect of pre-evaluating reductions. Note the decrease in operation count grows with the loop nest depth.}\end{figure}

It is then obvious that reaching an optimal form $\Lambda'$ is, in the general case, a challenging issue, since a full exploration of the expression tree transformation space is necessary. Even though we assumed we could generate $\Lambda'$ through a suitable transformation algorithm, the following issues should be addressed
\begin{itemize}
\item as opposed to what happens with hoisting in perfect multilinear loop nests, the temporary variable size would be proportional to the number of non-reduction loops crossed (in the example of Figure~\ref{code:loopnest}, $N \cdot O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L \cdot N \cdot O$ for those depending on $[L_e, L_i, L_j, L_k] $). This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime
\item the transformations exposing sub-expressions amenable to pre-evaluation could increase the arithmetic complexity within $L_k$ (e.g., expansion may increase the operation count; further examples will be provided later). This could overwhelm the gain inherent in pre-evaluation.
\end{itemize}

Before attacking this problem in Section~\ref{sec:optimal-synthesis}, we shall conclude our construction refining Definition~\ref{def:mln-optimality} for generic loop nests

\begin{Def}[Inner optimality of a loop nest]
\label{def:optimality}
The synthesis of a loop nest is optimal if, under a set of memory constraints \texttt{C}, the total amount of operations performed in all innermost loops is minimum.
\end{Def}

Note how the definition contemplates the possibility for a nest to have multiple innermost loops. In fact, multiple sub-nests could be rooted in the outermost loop, either because part of the input or as a result of some transformations (e.g., see Figure~\ref{code:loopnest_nored}).

\section{Optimal-form Loop Nests in Finite Element Integration}
\label{sec:optimal-synthesis}

In this section, we instantiate Definition~\ref{def:optimality} to our domain of interest, finite element integration. This will require reasoning at two different levels of abstraction: the math, in terms of the multilinear forms arising from the weak variational formulation of a problem; and the (partly multilinear) loop nests implementing such forms.

Our point of departure is the example loop nest in Figure~\ref{code:loopnest}. This loop nest is actually a simplified view of a typical bilinear form implementation. $L_e$ represents iteration over the elements of a mesh; $L_i$ derives from using numerical quadrature; $[L_j, L_k]$ implement the computation over test and trial functions. We deliberately omitted useless portions of code to not hinder readability and to avoid tying our discussion to specific forms.

We make the following observations. 1) $L \gg M, N, O$; that is, the number of elements $L$ is typically order of magnitude larger than both quadrature points ($M$) and degrees of freedom ($N$ and $O$ for test and trial functions); 2) $[L_j, L_k]$ (or simply $L_j$ with a linear form) is perfect and multilinear; this naturally descends from the translation of Equation~\ref{eq:quadrature} into a loop nest.

>>>>

\section{Problem Characterization}
\label{sec:optimal-impl}

In this section, we characterize optimality, as well as the transformation space that needs be explored to achieve it, for finite element integration.

\subsection{Loop Nests and Optimality}
In order to make the document self-contained, we start with reviewing basic compiler terminology.

@@

By mapping mathematical properties to the loop nest level, we introduce the
concepts of \textit{linear loop} and, more generally, (perfect) multilinear loop nest.

@@

\begin{Def}[Perfect multilinear loop nest]
A perfect multilinear loop nest of arity $n$ is a perfect nest composed of $n$ loops, in which all of the expressions appearing in the body of the innermost loop are linear in each loop $L_i$ separately.
\end{Def}

Since our focus is on finite element integration, for simplicity we restrict ourselves to the following, notable subclass of perfect multilinear loop nests.

\begin{Def}[Tensor-product loop nest]
A tensor-product loop nest is a perfect multilinear loop nest $\Lambda = [L_{i_{0}}, L_{i_{1}}, ..., L_{i_{n-1}}]$ in which all of the expressions appearing in the body of $L_{i_{n-1}}$ are (summations of) tensor products $t_{i_{0}, i_{1}, ..., i_{n-1}} = a_{i_{0}}b_{i_1}...z_{i_{n-1}}$.
\end{Def}

As shown later, tensor-product loop nests are important because they possess properties that simplify the synthesis of optimal code. They arise naturally when translating a multilinear form into code. Consider Equation~\ref{eq:quadrature} and the loop nest implementing it illustrated in Figure~\ref{code:loopnest}. The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises the element loop $L_e$, the integration loop $L_i$ (a reduction loop), and the tensor-product loop-nest $[L_j, L_k]$ over test and trial functions. The expression \texttt{F} implements the operator.

% (the Laplacian operator in the case of Equation~\ref{eq:quadrature}, but our discussion applies in general).
\begin{figure}[h]\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest.code}
\end{CenteredBox}\caption{The typical loop nest implementing a bilinear form.}\label{code:loopnest}\end{figure}

Our aim is to formulate a strategy to synthesize optimal implementations of such loop nests. In particular, we characterize optimality as follows.

\begin{Def}[Optimality of a loop nest]
\label{def:mln-optimality}
The synthesis of a loop nest is optimal if the number of operations performed in the loop nest is minimal.
\end{Def}

Note that Definition~\ref{def:mln-optimality} does not take into account memory requirements. If the loop nest were memory-bound -- the ratio of operations to bytes transferred from memory to the CPU being too low -- then speaking of optimality would clearly make no sense. Henceforth we assume to operate in a CPU-bound regime, in which arithmetic-intensive expressions need be evaluated. In the context of finite element, this is often true for more complex multilinear forms and/or higher order elements.

\subsection{Transformation Space}
To synthesize optimal implementations we need:
\begin{itemize}
\item a characterization of the transformation space for the class of loop nests considered
\item a cost model to select the optimal point in the transformation space
\end{itemize}
In this section, we construct the transformation space. We leave a few loose hands, which we progressively tie in Section~\ref{sec:optimal-synthesis}.

We start with introducing the fundamental notion of sharing.

@@

Figure~\ref{code:multi_loopnest_a} shows an example of a trivial tensor-product loop nest of arity $n=2$ with sharing along dimension $j$.

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 7pt \subfigure[With sharing]{\label{code:multi_loopnest_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}}}
~~~~~~~~~~~~~~
{\subfigcapskip = 2pt \subfigure[Optimal form]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}}}
\end{CenteredBox}\caption{A simple tensor-product loop nest}\end{figure}

We prove that tensor-product loop nests can be reduced to optimal form (i.e., the number of operations in the resulting loop nest is minimal) by eliminating sharing through a suitable algorithm.

\begin{Prop}
\label{prop:optimal-mln}
Assume $\Lambda = [L_{i_{0}}, L_{i_{1}}, ..., L_{i_{n-1}}]$ is a tensor-product loop nest with the body of $L_{i_{n-1}}$ being an independent basic block. Then we can determine an optimal $\Lambda'$ applying a sequence of steps aiming at eliminating sharing.
\end{Prop}
\begin{proof}
The demonstration is by construction and exploits linearity (recall that tensor-product loop nests are multilinear by definition). We start with ``flattening'' the expressions appearing in the body of $L_{i_{n-1}}$; that is, products are recursively expanded. This exposes the space of all possible factorizations. The terms appearing in the expanded expressions are tensor products of the form $a_{i_{0}}b_{i_1}...z_{i_{n-1}}$ (a tensor product involves $n$ arrays , each array indexed by a linear loop, or simply symbols).

We logically group the symbols into $n$ disjoint sets $S_i$, each $S_i$ containing the symbols indexed through $L_i$. These sets are sorted in ascending order based on their cardinality. Based on this, we create the factorization list: the ordered list of all factorizable symbols. We then progressively try to factorize these symbols. That is, we start factorizing along the loop that has the least number of unique symbols, and then we go on with the other loops.

Due to linearity, each factored product only has one symbol depending on $L_i$, and this symbol is now unique in the expression. The other symbols in the term, independent of $L_i$, are, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided. This procedure is always semantically correct, regardless of the symbols (loops) being considered: multilinearity ensures deterministic factorization; perfectness ensures safeness of hoisting.

Two observations related to the aforementioned code motion: 1) hoisting might require the creation of ``clone loops'' outside of $\Lambda$ to honor the potential dependency on loop indices; 2) hoisted expressions are replaced with temporaries, which need be added to a suitable set $S_i$ of factorizable symbols .

It remains to clarify why the chosen factorization ordering leads to optimality. Naturally, as we factorize a symbol, we are potentially pruning other factorization opportunities. If that is the case, we might wonder whether there was a better factorization strategy.  . Therefore, this is not possible (in the best case, the resulting factorization would be identical).

%The key observation is that in a tensor-product loop nest, all terms depend on $\Lambda$ through $n$ distinct symbols.

Same cardinality, try both
\end{proof}

%\begin{proof}
%By construction. Loop-dependent terms are logically grouped into $n$
%disjoint sets $S_i$, each $S_i$ containing all terms depending on
%$L_i$. These sets are sorted in descending order based on their
%cardinality. We produce a new loop permutation that reflects the ${S_i}$
%ordering; that is, loops are placed such that each loop has fewer unique
%terms than the outer loops containing it. Since the loop nest is perfect,
%any such permutation is semantically correct.  Starting from $L_{i_{n-1}}$, we can apply the sharing-removal procedure described in Proposition~\ref{prop:share-removal}. This renders $L_{i_{n-1}}$ optimal. In particular, the number of operations is equal to $\Lambda_{ops} = \# S_{i_{n-1}} + (\# S_{i_{n-1}} - 1)$, in which the second term represents the cost of summing the $\# S_{i_{n-1}}$ terms.
%\end{proof}

For example, by applying this procedure to the code in Figure~\ref{code:multi_loopnest_a}, we obtain the optimal form in Figure~\ref{code:multi_loopnest_b}.

However, we observe that, for the larger class of finite element integration loop nests, the presence of sharing is a \textit{sufficient but not necessary} condition for being in \textit{non-optimal} form. Consider again the bilinear form implementation in Figure~\ref{code:loopnest}. We pose the following question: are we able to identify sub-expressions within $F$ for which the reduction imposed by $L_i$ can be pre-evaluated, thus obtaining a decrease in operation count proportional to the size of $L_i$, $M$? The transformation we look for is exemplified in Figures~\ref{code:loopnest_red} (input) and~\ref{code:loopnest_nored} (output); Figure~\ref{code:loopnest_red} can be seen as a simple instance of the abstract loop nest in Figure~\ref{code:loopnest}.

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 19pt \subfigure[With reduction]{\label{code:loopnest_red}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_red.code}}}
~~~~~~~~~~
{\subfigcapskip = 2pt \subfigure[Pre-evaluated reduction]{\label{code:loopnest_nored}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_nored.code}}}
\end{CenteredBox}\caption{Exposition (through factorization) and pre-evaluation of a reduction.}\end{figure}

Pre-evaluation opportunities could have to be exposed, for instance through an exploration of the expression tree transformation space, which can be challenging. Further, the following issues arise when considering pre-evaluation opportunities:
\begin{itemize}
\item as opposed to what happens with hoisting in tensor-product loop nests, the temporary variable size is proportional to the number of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $N \cdot O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L \cdot N \cdot O$ for those depending on $[L_e, L_i, L_j, L_k] $). This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime
\item the transformations exposing pre-evaluation opportunities could increase the arithmetic complexity (e.g., expansion may increase the operation count; further examples will be provided later). This could overwhelm the gain inherent in pre-evaluation.
\end{itemize}

To summarize, so far we have highlighted four problems:
\begin{enumerate}
\item The need for an algorithm to expose pre-evaluation opportunities
\item Potential explosion in working set size
\item Potential increase in operation count due to manipulation of the expression tree
\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities
\end{enumerate}

We tackle these points in Section~\ref{sec:optimal-synthesis}. In the perspective of addressing point 2, it is useful to conclude this section refining our optimality statement as follows.

\begin{Def}[Optimality of a loop nest with bounded working set]
\label{def:optimality}
The synthesis of a loop nest is optimal if, under a set of memory constraints $C$, the number of operations performed in the loop nest is minimal.
\end{Def}

\section{Optimal Integration Routines}
\label{sec:optimal-synthesis}

In this section, we will reason at two different levels of abstraction: the math, in terms of the multilinear forms arising from the weak variational formulation of a problem; and the (partly multilinear) loop nests implementing such forms.

Our point of departure is the example loop nest in Figure~\ref{code:loopnest}. We make the following observations. 1) $L \gg M, N, O$; that is, the number of elements $L$ is typically order of magnitude larger than both quadrature points ($M$) and degrees of freedom ($N$ and $O$ for test and trial functions); 2) $[L_j, L_k]$ (or simply $L_j$ with a linear form) is perfect and multilinear; this naturally descends from the translation of Equation~\ref{eq:quadrature} into a loop nest.
